Understanding the impacts of video-guided activities on parent-child interaction

### 2021-01-08 paper/manuscript.Rmd
abstract, limitations

abstract: |
  Early parenting practices play an important role in shaping children’s future outcomes. 
  In particular, high-quality early interactions can facilitate language learning and school performance. 
  The rise of phone-based parenting applications ("apps") could deliver low-cost interventions on parenting style to a wide variety of populations, especially the parents of very young children, who are often difficult to reach in other ways. 
  Yet little is known about the effects of communicating to parents through app-based interventions. 
  In two studies (one preregistered), we showed parents short videos depicting age-appropriate parent-child activities from a parenting app. 
  We found that after watching the video, parents spoke more and made more bids for joint attention, as compared with controls who watched no video (experiment 1) or a science video (experiment 2). 
  These results suggest that activity videos can lead to positive changes in parent engagement, providing support for the use of such videos in parenting interventions.
keywords          : "language development; parent intervention; childhood development; joint attention; lexical diversity"
wordcount         : "5066"
bibliography      : ["library.bib"]
floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
header-includes:
  - \usepackage{float}
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

@@

# Research Highlights

* After watching short activity videos in a parenting app, parents made more attempts to engage with their children and spoke more to them.
* Older toddlers showed greater engagement after parents watched the activity videos, relative to younger toddlers.
* These activity videos led parents to set more goals for their child during play. 
* Our findings suggest that parent-led activities guided by apps may be helpful in scaffolding the play of older toddlers.

@@

The quantity and quality of early language input has been found to be strongly associated with later language and academic outcomes [@Hart1995; @Marchman2008; @Cartmill2013; @HirshPasek2015; see @Schwab2016 for a review]. Thus, because of the potential for large downstream effects [@Heckman2006], there is tremendous interest in interventions that change children's language environment. 
And because parents define a large portion of that environment, especially before the onset of formal schooling, parent behavior is a critical locus for such interventions. 
Many effective parenting interventions require large resource investments and require many hours of in-person contact [@Jamaica2014; @PerryPreschool2004; @Leung2018], making implementation at scale a daunting proposition. 
For this reason, many researchers targeting early language are interested in delivering parenting interventions remotely -- through texts, apps, and videos delivered on digital devices. 
But what do parents take away from these short messages about what to do with or how to talk with their children?

The content provided by digital parenting interventions runs the gamut from general parenting messages and facts from child development research to specific advice, coaching, and suggested activities. 
A growing body of evidence suggests that these digital interventions can be effective across a range of cultures, income levels, and children's ages [for a review, see @Breitenstein2014].
For example, in contrast to a face-to-face parent training intervention, a tablet-based version saw significantly higher session completion rates (51% attendance vs. 85% module completion) and comparable or larger effect sizes on parents' and children's (aged 2 to 5 years) behavior [@Breitenstein2016].
Often, however, the theory of change presupposed by such interventions is relatively vague.
Both within and outside the realm of academic interventions, messages to parents of young children often seek to provide knowledge about some aspect of development (e.g., early language), often in tandem with a suggestion regarding activities.
Such messages are assumed to inform parents' choice of behaviors, spurring them to engage in some target activity, which is assumed to be more stimulating than what parents would have done otherwise.

@@

But is this theory of change correct? That is, does the provision of knowledge and activities lead to higher-quality play? 
Alternatively, by focusing parents on a specific activity, this approach could be flawed, causing parents to over-focus on achieving the superficial goals of the activity. 
This problem might be especially likely with video messages, which could encourage parents to try to mimic a model's specific speech and/or actions.
Attempting to reproduce such surface details of a video-guided activity could in turn result in less high-quality talk, with less responsiveness to their child's play.
Another possibility is that these messages might produce the desired effect, but only for those parents who already have a general orientation towards children's early learning.

Our current experiments were designed to make a direct test of this question: How do parents change their interactions with young children on the basis of short video parenting messages? 
In two experiments, we collected data from parent-child dyads in a local children's museum. 
We showed parents in the experimental group a single short video modeling an interactive toy-based activity along with a scientific justification. 
Parents in the control group received either no video (Experiment 1) or a video of a recent finding in developmental psychology (Experiment 2). 
We then gave the toys from the video to all dyads and videotaped their interactions, coding for caregivers' language quantity and quality as well as joint attention.

@@

We were interested in how digital parenting advice alters parents' interactions with their children. 
We specifically set out to test whether activity suggestions led to higher-quality play, a presupposition of many early parenting interventions. 
Our experiments explored this question by randomly assigning parents to different advice conditions and then observing their behavior in short free-play sessions, with quality of play assessed through measures of parent language and joint attention. 
In two experiments, we found that activity videos increased the rate of parents' bids for joint attention as compared with no video (Experiment 1) and a comparable science video (Experiment 2). 
In some cases---especially in older children---these bids were successful in increasing engagement. 
We also observed differences in parents' talk that were broadly similar across both experiments, with a greater quantity of language but a similar breadth of vocabulary (leading to lower measures of lexical diversity). 
Exploratory corpus analysis identified the words most characteristic of the activity videos as being related to requests (e.g., "give", "put", "take") whereas control conditions featured more invitational words (e.g., "say", "let", "like"), often asking about animal noises ("What does the X say?").

@@

Our study has a number of limitations related to design and sample, each of which suggests possible future directions. 
First, our design was intentionally short and minimal; future studies should investigate whether changes in parents' speech and attempts to engage their children could persist across a longer timespan (perhaps with a broader set of activities being provisioned). 
A longer-term study would also address whether consistent increases in parent bids would lead children to respond by engaging more with their parent.
Second, our design assumes that parents have access to the materials needed to complete the suggested activities; this assumption may be unrealistic for any parent, but especially for the parents who are most likely to be targeted for early parenting interventions. 
Providing materials may be critical for the success of activity suggestions. 
Finally, our sample is a convenience sample drawn from a museum, but it skews towards higher socio-economic status households as well as those families who are well-disposed towards visiting a museum (perhaps because they value education) and are interested in participating in research. 
A key goal for future research is to assess the generality of these findings across populations.

@@

In sum, the results of this study show that digital parenting videos recommending play activities can lead to short-term increases in parents' attempts to engage their young children, both verbally and non-verbally.

@@

This work was supported by a gift and a contract from Kinedu, Inc. to the Language and Cognition Lab. 
Thanks to Samaher Radwan and Megan Merrick for behavioral coding of the videos, and to members of the Language and Cognition Lab at Stanford for helpful discussion.

>>>>

abstract: |
  Early parenting practices play an important role in shaping children’s future outcomes. 
  In particular, high-quality early interactions can facilitate language learning and school performance. 
  The rise of phone-based parenting applications ("apps") could deliver low-cost interventions on parenting style to a wide variety of populations, especially the parents of very young children, who are often difficult to reach in other ways. 
  Yet relatively little is known about the local effects of communicating to parents about how they should interact via videos. 
  In two studies (one preregistered), we investigated effects of short videos on parent behavior.
  We showed parents videos depicting age-appropriate parent-child activities from a parenting app. 
  We found that after watching the video, parents spoke more and made more bids for joint attention, as compared with controls who watched no video (experiment 1) or a science video (experiment 2). 
  These results suggest that activity videos can lead to local changes in parent engagement, providing support for the use of such videos as a part of broader app-based parenting interventions.
keywords          : "language development; parent intervention; childhood development; joint attention; lexical diversity"
wordcount         : "5066"
bibliography      : ["library.bib"]
floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
header-includes:
  - \usepackage{float}
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

@@

@@

The quantity and quality of early language input has been found to be strongly associated with later language and academic outcomes [@Hart1995; @Marchman2008; @Cartmill2013; @HirshPasek2015; see @Schwab2016 for a review]. Thus, because of the potential for large downstream effects [@Heckman2006], there is tremendous interest in interventions that change children's language environment. 
And because parents define a large portion of that environment, especially before the onset of formal schooling, parent behavior is a critical locus for such interventions. 
Many effective parenting interventions require large resource investments and require many hours of in-person contact [@Jamaica2014; @PerryPreschool2004; @Leung2018], making implementation at scale a daunting proposition. 
For this reason, many researchers targeting early language are interested in delivering parenting interventions remotely -- through texts, apps, and videos delivered on digital devices. 
But what do parents take away from these short messages about what to do with or how to talk with their children? 
In this paper, rather than conducting a large-scale intervention, we investigate the ways that parents change their local behavior with children immediately after seeing a short video instructing them on an activity, investigating one of the building blocks of many larger interventions.

The content provided by digital parenting interventions runs the gamut from general parenting messages and facts from child development research to specific advice, coaching, and suggested activities. 
A growing body of evidence suggests that these digital interventions can be effective across a range of cultures, income levels, and children's ages [for a review, see @Breitenstein2014].
For example, in contrast to a face-to-face parent training intervention, a tablet-based version saw significantly higher session completion rates (51% attendance vs. 85% module completion) and comparable or larger effect sizes on parents' and children's (aged 2 to 5 years) behavior [@Breitenstein2016].
Often, however, the theory of change presupposed by such interventions is relatively vague.
Both within and outside the realm of academic interventions, messages to parents of young children often seek to provide knowledge about some aspect of development (e.g., early language), often in tandem with a suggestion regarding activities.
Such messages are assumed to inform parents' choice of behaviors, spurring them to engage in some target activity, which in turn is assumed to be more stimulating than what parents would have done otherwise.

@@

But is this theory of change correct in the case of digital parenting interventions? 
That is, does the provision of knowledge and activities via short videos or messages actually lead to higher-quality play? 
Alternatively, this approach could be flawed. 
For example, focusing parents on a specific activity could cause them parents to over-focus on achieving the superficial goals of the activity while neglecting the broader goals of fostering more -- and more responsive -- interaction. 
This problem might be especially likely with video messages, which could encourage parents to try to mimic a model's specific speech and/or actions.
Attempting to reproduce the surface details of a video-guided activity could in turn result in less high-quality talk, with less responsiveness to their child's play.
(Speaking as a parent, one author of the current paper has found himself occasionally so engrossed in attempting to execute an activity with his children that he has lost track of whether the children are continuing to enjoy it!).

Thus, our goal was to gather information about whether the use of short video parenting messages as a component of broader interventions is supported by their local effects on parent-child interaction and language use. 
Our current studies were designed to assess how parents change their interactions with young children on the basis of these videos.  
We made a first step towards assessing this question by performing two studies within the context of a convenience sample of parent-child dyads in a local children's museum. 
We showed parents in the experimental group a single short video modeling an interactive toy-based activity along with a scientific justification. 
Parents in the control group received either no video (Experiment 1) or a video of a recent finding in developmental psychology (Experiment 2). 
We then gave the toys from the video to all dyads and videotaped their interactions, coding for caregivers' language quantity and quality as well as joint attention.

@@

Activity suggestions, often in the form of videos, are a component of many larger digital interventions for parents. 
We were interested in how these short activity videos alter parents' interactions with their children -- in particular, whether they actually lead to higher quality language or interaction, a presupposition of their inclusion in broader interventions. 
Towards this goal, we performed two short randomized experiments with a convenience sample of parent-child dyads. 
We observed their behavior in short free-play sessions after viewing either activity videos or control videos, with quality of play assessed through measures of parent language and joint attention. 
In two experiments, we found that activity videos increased the rate of parents' bids for joint attention as compared with no video (Experiment 1) and a comparable science video (Experiment 2). 
In some cases---especially in older children---these bids were successful in increasing engagement. 
We also observed differences in parents' talk that were broadly similar across both experiments, with a greater quantity of language but a similar breadth of vocabulary (leading to lower measures of lexical diversity). 
Exploratory corpus analysis identified the words most characteristic of the activity videos as being related to requests (e.g., "give", "put", "take") whereas control conditions featured more invitational words (e.g., "say", "let", "like"), often asking about animal noises ("What does the X say?").

@@

Our study is a first step towards testing one component of a broader theory of change, namely that provisioning activities to parents can lead to more and more positive parenting interactions. 
But even given our results, there are of course myriad reasons why digital parenting interventions might not be successful.
Even with the best of intentions, parents viewing activities may not actually carry those activities out with greater frequency; other constraints of time and resources might be more determinative of parent behavior. 
Our results begin to address only one baseline concern about such interventions -- one that the authors held prior to performing these studies -- namely that watching these videos might simply not result in any measurable changes to parent behavior. 
That particular worry appears unfounded, though whether the changes we observed in parent behavior can be harnessed in a broader intervention remains a question for future work.

Our study has a number of further limitations related to design and sample, each of which suggests possible future directions. 
First, our design was intentionally short and minimal; future studies should investigate whether changes in parents' speech and attempts to engage their children could persist across a longer timespan (perhaps with a broader set of activities being provisioned). 
A longer-term study would also address whether consistent increases in parent bids would lead children to respond by engaging more with their parent.
Second, our design assumes that parents have access to the materials needed to complete the suggested activities; this assumption may be unrealistic for any parent, but especially for the parents who are most likely to be targeted for early parenting interventions. 
Providing materials may be critical for the success of activity suggestions. 
Third, our study evaluated only one source of videos and did not systematically identify those characteristics of the videos that might lead to success.
Finally, our sample is a convenience sample drawn from a museum, but it skews towards higher socio-economic status households as well as those families who are well-disposed towards visiting a museum (perhaps because they value education) and are interested in participating in research. 
A key goal for future research is to assess the generality of these findings across populations.

@@

In sum, our findings suggest that digital parenting videos recommending play activities can lead to short-term increases in parents' attempts to engage their young children, both verbally and non-verbally. We hope that future work on this topic attempts to realize this promise in the context of broader intervention research.

@@

This work was supported by both an unrestricted gift and a research contract from Kinedu, Inc. to the Language and Cognition Lab. 
Thanks to Samaher Radwan and Megan Merrick for behavioral coding of the videos, and to members of the Language and Cognition Lab at Stanford for helpful discussion.
